### 2010-03-05 docs/algdraft/MaterializeDBV1.tex
Update the documents.

\newtheorem{example}{Example}[section]
\newtheorem{definition}{Definition}[section]
\newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}
\begin{document}

\section{Key constraints}

%motivation to have key constraints.
In real world, the dataset collected by an ecologist to describe
measurents and observations is not normalized. I.e., two different
rows in the table may have a big amount of ``redundant'' 'information
to describe an object (e.g., entity).

As discussed in \cite{***} %ER paper, 
we can use the OBOE cenceptual model to capture the semantic
relationships from the columns to their real meaning. 
However, the mapping from the columns to the OBOE model still cannot
directly solve the problem of getting concise informatin (e.g., unique
entity instances, unique observation instances, etc.)

Given Table \ref{tb:pltarea}(a), it records the fact that four times
of dbh of some trees are taken. 
Very often, an ecologist may ask questions like this:  Give me the
average dbh for every different trees. 
First, we have four rows here for trees. 
But how many tree entity instances are here is a question.

There are several cases to consider:
\begin{itemize}
\item Case 1: The naive extreme way to interpret the data is that each
  observation is from different tree entity. Then, we have {\bf four}
  tree entities. This may be too strict. People may say, well, I have
  some observations for the same entity. 
\item Case 2: The second naive extreme way is to interpret that different {\em spp} represent the different tree entity. 
That's obvious that {\em piru} is different from {\em abba}. 
With this constraint, we get {\bf two} tree entities. However, an
obvious fact is also true that one {\em piru} tree cannot at the
different plots. (It is another story if this tree was migrated from one plot to another.) 
\item Case 3: Following Case 2, people want to further limit that the
  same {\em spp} in the same {\em plt} should represent the same tree
  entity set.  
%To achieve this, we use {\em identifying yes}.
%   Now, it should return
%	\[(A, piru, 36), (B, piru, 33.2), (B, abba, 34)\]
\end{itemize}

%Q2: Why we need to distinguish the same entities using {\em key yes}
%and {\em distinct yes}) to identify the same observation? 
%For the example, we have one observation for splot {\em A}.
%What's the semantic purpose of this? What kind of query may need this? 
%E.g., how many spots in this dataset?

\begin{table}[htb]
\begin{center}
\begin{tabular}{cc}
\begin{tabular}{|l|l|l|}
\hline
plt & spp & dbh\\\hline
A & piru & 35.8 \\\hline
A & piru & 36.2 \\\hline
B & piru &33.2 \\\hline
B&abba&34\\\hline
\end{tabular}
&
\begin{tabular}{|l|l|l|l|}
\hline
plt & area & spp & dbh\\\hline
A & 1.0 & piru & 35.8 \\\hline
A & 1.1 & piru & 36.2 \\\hline
B & 2.0 &piru &33.2 \\\hline
B& 2.0 & abba&34\\\hline
\end{tabular}\\
(a) & (b)
\end{tabular}
\end{center}
\vspace{-0.2in}
\caption{Dataset}
\label{tb:pltarea}
\end{table}

Obviously, to answer this question, we not only need to figure out
what are the observations and measurements, we also need proper
constraints to help us understand whether different observations are
from the same entity or not.

To tackle this problem, we proposed to use {\em key yes}, {\em
  distinct yes} and {\em identifying yes} constraints.

\from{HP} {Describe or define {\em key yes},  {\em distinct yes},  {\em identifying yes}}

\begin{definition}[Key measurement types]{
The key mensurement tytpes of an observation are the measurement types
that are used to distinguish one observation/entity instance from the
other. }
\end{definition}
In most simple case, when an observation type does not have context, 
the key measurement types are specified with ``key yes''.

{\bf Key constraints}: in this report (paper),  we use the term {\em
  key constraints} to refer to the general constraints of the above
three types {\em key yes}, {\em distinct yes} and 
{\em identifying yes}.

\section{Valid annotation constraints}

Based on the semantic meaning of these key constraints, we can see
that they have some correlations. 
Before we give any further examples to explain such correlations, we
list the correlation rules among them.

\begin{itemize}
\item {\bf Rule 1}: If an observation type is specified with {\em distinct yes}, all
  its measurement types are automatically marked with {\em key yes}. 
\item {\bf Rule 2}: If the context of an observation type is specified with {\em identifying
    yes}. This observation type MUST have some key measurement
  types. And the context observation type also MUST have key
  measurement types too. 
\end{itemize}

The following example shows an example why Rule 1 is needed. 
Given the data in Table \ref{tb:pltarea}(b) and the following
annotation.

\noindent
{\bf observation} ``o1''  \textcolor{blue}{distinct yes}\\
\verb|    |{\bf  entity} ``Plot''\\
\verb|    |{\bf measurement} ``m1'' \textcolor{blue}{key yes}\\
\verb|        | {\bf characteristic} ``EntityName'' \\
\verb|        | {\bf standard} ``Nominal''\\
\verb|    | {\bf measurement} ``m2''\\
\verb|        | {\bf characteristic} ``area'' \\
\verb|        | {\bf standard} ``sqft''\\
\verb|        |$\cdots$

According to the annotation, if two plots have the same value for
``EntityName'', they represent the same plot observation. 
Obviously, it has problem to interprete the data in Table
\ref{tb:pltarea} with this annotatioin. 
E.g., the first and second rows catch information about plot with
EntityName $A$. Accoring to the annotation, they should be the same
plot obervation. 
However, the data shows that this one plot has two different areas
$1.0$ and $1.1$, so, there is confusion here.

For this case, we can have several ways to make the annoation
consistent. 
The first alternative to tackle this confusion is to change the
annotation.  We can let $m_2$ automatically becomes key measurement type of
$o_1$. Then, every different ($m_1$, $m_2$) value pair denote a
different observation instance which come from the different entity
instance.
The second alternative is to get rid of {\em distinct yes} for $o_1$,
which means that $(A,1.0)$ and $(A,1.1)$ are different observation instances of the same entity instance
$A$.

%Here are some very basic rules to restrict the correlations among these constraints. 
%\begin{itemize}
%\item ``Distinct yes'' on an observation type implies that all its measurements collectively form a key of this observation type. 
%In the above example, we can specify ``distinct yes'' on ``$o_1$''
%without {\em explicitly} denote ``key yes'' on the measurements $m_1$
%and $m_2$. 
%The semantic meaning is that ($m_1$, $m_2$) collectively form key measurement for $o_1$.  
%\item Logically, we can specify ``Distinct no'' on an observation type and specify ``key yes'' on all of its measurements.
%In this case, when two rows (for the observation type) have the same
%values for these key measurements, they represent the different
%observation instances. Such a case means that there may exist other
%``variables'' that can distinguish one observation instance from the
%other, but this dataset does not catch this variable ({\bf HP: is
%``variable'' the right word to use? }). So, in real application, we
%do not encourage such situation to appear. 
%\end{itemize}

In our materialization algorithm, we assume that the annotations on the dataset are valid, i.e., they comply with the rules among these constraints.

@@

The output of the algorithm is a set of materialized tables represented in the OBOE model and denoted by $OBOE.*$. In detail,
\begin{itemize}
\item $OBOE.Observation=\{\langle \underline{ObsId}, ObsType, EntId \rangle\}$ keeps all the observation instances;
\item $OBOE.Measurement=\{\langle \underline{MeasId}, ObsId, MeasType, Val\rangle\}$ for all the measurement instances;
\item $OBOE.Entity=\{\langle \underline{EntId}, EntType\rangle\}$ for all the entity instances;
\item $OBOE.Context =\{\langle \underline{ObsId, ConextObsId, ContextType}\rangle\}$ for all the context instances;
\end{itemize}

>>>>

\newtheorem{example}{Example}[section]
\begin{document}

\section{Valid annotation constraints}
Among the {\em key}, {\em distinct}, and {\em identifying} constraints,
there are correlations between ``key yes'' and ``distinct yes''.

The following example gives us a hint of invalid annotations. Given the data in Table \ref{tb:pltarea}. The following annotation is {\em invalid}.

\noindent
{\bf observation} ``o1''  \textcolor{blue}{distinct yes}\\
\verb|    |{\bf  entity} ``Plot''\\
\verb|    |{\bf measurement} ``m1'' \textcolor{blue}{key yes}\\
\verb|        | {\bf characteristic} ``EntityName'' \\
\verb|        | {\bf standard} ``Nominal''\\
\verb|    | {\bf measurement} ``m2''\\
\verb|        | {\bf characteristic} ``area'' \\
\verb|        | {\bf standard} ``sqft''\\
\verb|       |$\cdots$ Not finished....
\\
According to the annotation, if two plots have the same value on EntityName, they represent the same plot observation. 
Obviously, it has probelem to interprete the data in Table \ref{tb:pltarea} with this annotatioin. 
E.g., the first and second row catching information about plot with EntityName $A$. Accoring to the annotation, they should be the same plot obervation.
However, the data shows that this one plot has two different areas $1.0$ and $1.1$, so, there is confusion here.
\\
\begin{table}[htb]
\begin{center}
%\begin{tabular}{cc}
%\begin{tabular}{|l|l|l|}
%\hline
%plt & spp & dbh\\\hline
%A & piru & 35.8 \\\hline
%A & piru & 36.2 \\\hline
%B & piru &33.2 \\\hline
%B&abba&34\\\hline
%\end{tabular}
%&
\begin{tabular}{|l|l|l|l|}
\hline
plt & area & spp & dbh\\\hline
A & 1.0 & piru & 35.8 \\\hline
A & 1.1 & piru & 36.2 \\\hline
B & 2.0 &piru &33.2 \\\hline
B& 2.0 & abba&34\\\hline
\end{tabular}\\
%(a) & (b)
%\end{tabular}
\end{center}
\vspace{-0.2in}
\caption{Dataset}
\label{tb:pltarea}
\end{table}

{\bf From HP:}  There are several ways to let the data comply with the annotaion without confusion. 
Maybe too detailed, I omit this here.

Here are some very basic rules to restrict the correlations among these constraints. 
\begin{itemize}
\item ``Distinct yes'' on an observation type implies that all its measurements collectively form a key of this observation type. 
In the above example, we can specify ``distinct yes'' on ``$o_1$'' without {\em explicitly} denote ``key yes'' on the measurements $m_1$ and $m_2$. 
The semantic meaning is that ($m_1$, $m_2$) collectively form key measurement for $o_1$.  
\item Logically, we can specify ``Distinct no'' on an observation type and specify ``key yes'' on all of its measurements.
In this case, when two rows (for the observation type) have the same values for these key measurements, they represent the different observation instances. Such a case means that there may exist other ``variables'' that can distinguish one observation instance from the other, but this dataset does not catch this variable ({\bf HP: is ``variable'' the right word to use? }). So, in real application, we do not encourage such situation to appear. 
\end{itemize}

In our algorithm, we assume that the annotations on the dataset are valid, i.e., they comply with the rules among these constraints.

{\bf From HP:} How detailed to discuss this?

@@

The output of the algorithm is a set of materialized tables represented in the OBOE model and denoted by $OBOE.*$. In detail,
\begin{itemize}
\item $OBOE.Observation=\{\langle \underline{ObsId}, EntId \rangle\}$ keeps all the observation instances;
\item $OBOE.Measurement=\{\langle \underline{MeasId}, ObsId, MeasType, Val\rangle\}$ for all the measurement instances;
\item $OBOE.Entity=\{\langle \underline{EntId}, EntType\rangle\}$ for all the entity instances;
\item $OBOE.Context =\{\langle \underline{ObsId, ConextObsId, ContextType}\rangle\}$ for all the context instances;
\end{itemize}

### 2010-03-05 docs/algdraft/UseCaseKey.tex
Update the documents.

\documentclass[10pt]{article}
%\title{}
%\author{}
%\date{}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{a4}
\usepackage{color}
\renewcommand\floatpagefraction{0.99}
\renewcommand\topfraction{0.99}
\renewcommand\bottomfraction{0.99}
\renewcommand\textfraction{.05}
\setcounter{totalnumber}{5}
%\setlength{\parindent}{0pt}
%\pdfpagewidth 8.5in
%\pdfpageheight 11in
\newtheorem{example}{Example}[section]
\begin{document}

{\bf Use cases to show that it's needed to have {\em key yes}, {\em identifying yes} and {distinct yes}}

Q1: Why we need to distinguish the same entities using {\em key yes} and {\em identifying yes})?

Assume the following tabe is the measurement for some plant tree at different spots.

Consider this question that a user may ask. Give me the average dbh for every {\em piru} tree (i.e., tree entity). 
First, we have three observations here.
But how many tree entities here is a question.

There are several cases to consider:
\begin{itemize}
\item Case 1: The naive extreme way to interpret the data is that each observation is from different tree entity. Then, we have {\bf four} tree entities. 
	This may be too strict. People may say, well, I have some observations for the same entity. 
\item Case 2: The second naive extreme way is to interpret that different {\em spp} represent the different tree entity. 
That's obvious that {\em piru} is different from {\em abba}. 
With this constraint, we get {\bf two} tree entities. 
\item Case 3: the assumption of case two has some obvious problem. 
People want to further limit that the same {\em spp} in the same {\em plt} should represent the same tree entity set. 
To achieve this, we use {\em identifying yes}.
   Now, it should return
	\[(A, piru, 36), (B, piru, 33.2), (B, abba, 34)\]
\end{itemize}

\begin{table}[htb]
\begin{center}
\begin{tabular}{cc}
\begin{tabular}{|l|l|l|}
\hline
plt & spp & dbh\\\hline
A & piru & 35.8 \\\hline
A & piru & 36.2 \\\hline
B & piru &33.2 \\\hline
B&abba&34\\\hline
\end{tabular}
&
\begin{tabular}{|l|l|l|l|}
\hline
plt & area & spp & dbh\\\hline
A & 1.0 & piru & 35.8 \\\hline
A & 1.1 & piru & 36.2 \\\hline
B & &piru &33.2 \\\hline
B& & abba&34\\\hline
\end{tabular}\\
(a) & (b)
\end{tabular}
\end{center}
\vspace{-0.2in}
\caption{Dataset}
\end{table}

Q2: Why we need to distinguish the same entities using {\em key yes} and {\em distinct yes}) to identify the same observation? 
For the example, we have one observation for splot {\em A}.
What's the semantic purpose of this? What kind of query may need this? 
E.g., how many spots in this dataset?

{\bf Note 1}: if one observation type is marked with {\em distinct yes}, all its measurements should be marked with {\em key yes}. 
Otherwise, we may have the same observation with different measurement values. 
E.g., what will happen for the following:

\noindent
{\bf observation} ``o1''  \textcolor{blue}{distinct yes}\\
\verb|    |{\bf  entity} ``Plot''\\
\verb|    |{\bf measurement} ``m1'' \textcolor{blue}{key yes}\\
\verb|        | {\bf characteristic} ``EntityName'' \\
\verb|        | {\bf standard} ``Nominal''\\
\verb|    | {\bf measurement} ``m2''\\
\verb|        | {\bf characteristic} ``area'' \\
\verb|        | {\bf standard} ``sqft''\\

Will $(A, 1.0)$ and $(A,1,1)$ be treated as the same observation? 
According to the semantic meaning, they are the same observation because they have the same value on the key measurement and this observation type is marked witht {\em distinct yes}. 
However, there is something wrong here.

\begin{table}[htb]
\begin{center}

\end{center}
\vspace{-0.2in}
\caption{Dataset}
\end{table}

Based on this note, it seems like it is not useful to denote {\em distinct yes}.
Basically, once all the measurements are marked with {\em key yes}, it automatically infers that it is distinct yes.

>>>>

\documentclass[10pt]{article}
%\title{}
%\author{}
%\date{}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{a4}
\usepackage{color}
\renewcommand\floatpagefraction{0.99}
\renewcommand\topfraction{0.99}
\renewcommand\bottomfraction{0.99}
\renewcommand\textfraction{.05}
\setcounter{totalnumber}{5}
%\setlength{\parindent}{0pt}
%\pdfpagewidth 8.5in
%\pdfpageheight 11in
\newtheorem{example}{Example}[section]
\newcommand{\from}[2]{{\bf[{\sc from #1:} #2]}}
\begin{document}

How to annotate the above dataset using ``key yes'', ``distinct yes''
and ``identifying yes''.

\begin{table}
\begin{center}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
Plot & sub-plot & Tmnt & Sp & Ind &wt\\\hline
1 & A & X &Aus&1&10\\\hline
1 & A & C &Bus&1&20\\\hline
1 & B & X &Aus&3&10\\\hline
1 & B & C &Bus&4&10\\\hline
2 & A & X &Aus&1&20\\\hline
2 & A & C &Bus&4&10\\\hline
2 & B & X &Aus&5&20\\\hline
2 & B & C &Bus&4&10\\\hline
\end{tabular}
\end{center}
\caption{A dataset with more complex information}
\label{tb:complexdb}
\end{table}

\noindent  Given the dataset in Table \ref{tb:complexdb}, users have different situations to catch. 
\begin{itemize}
\item Requirement 1: {\em Plot} with label ``1'' should refer to the same one physical plot (i.e., the Plot in the first 4 row means the same thing); similarly, 
  {\em plot} with label ``2'' should refer to the second physical plot (i.e., the Plot in the last 4 row means the same thing). 
\begin{itemize} 
\item This can be captured in annotation by putting ``Distinct yes'' for observation type {\em Plot} and ``key yes'' for its measurement type {\em PlotLabel}. 
\end{itemize}
\item Requirement 2: {\em sub-plot}s with the same lable should refer to the same physical sub-plot if they are within the same plot; 
but the sub-plot with the same label with different {\em Plot} label are different sub-plots.
E.g., Row 1 \{Plot=1, sub-plot=A\} refers the same sub-plot as that in Row 2, but is different from the one in Row 5 \{Plot=2, sub-plot=A\}. 
\begin{itemize} 
\item This can be captured by putting ``Distinct yes'' for observation type {\em SubPlot} , ``key yes'' for its measurement type {\em SubPlotLabel}. We need to denote {\em Plot} is its context and with {\em identifying yes} specified on this context. 
\end{itemize}
\item Requirement 3: {\em Tmnt} with the same lable should refer to the same treatment process (So that we can aggregate on different treatment process, e.g., on ``X'' or on ``C''.)  But the treatment at different sub-plot should refer to different treatment. 
 \begin{itemize} 
\item The first requirement can  be captured by treating all the
  Treatment with value ``X'' as the different entity instances with
  the same type (TmntType). The second requirement can be captured by
  treating the treatments in different sub-plots as observations of
  the different entity instances.  I.e., treatments in row 1 and row 3
  are of but are different observation instances which are of
  different entity instances. 
\item {\em Tmnt} has the context {\em sub-plot}. 
We don't need to specify ``key yes'', ``distinct yes'', or ``identifying yes''. 
%At the first glance, to represent this, {\em TmntType} for {\em Tmnt} should be specified with ``key yes''. 
%It should have context {\em sub-plot} which is specified with ``identifying yes''. 
%\item After further analysis, {\bf  one question arises: 
%The key measurements for the treatment observation is different from the key measurement  of the entity treatment}. 
%After considering the context, the key measurements for the treatment observation are \{Plotlabel, SubPlotLabel, TmntType\}.  
%When two rows have the same value on these measurements, they represent the same observation instance.
%However, the key measurement for the treatment entity is just \{TmntType\}. When two rows have the same value on it, they represent %the same entity instance. 
%The {\em identifying} constraint can only capture the observation context. 
%{\bf This problem is more obvious when we analyze Case 5}. 
%\item Another different annotation may be applied to catch this semantic. E.g., treat the {\em treatments} in different rows as different entity instances. 
%This way, the observation type and the entity type have the same key measurement types \{Plotlabel, SubPlotLabel, TmntType\}. 
%However, this problem still exists for Case 4 and Case 5.

Summarization questions: \\
(1) Give me the average weight of the individuals with treatment ``X''. 
How can this question be answered after the annotation and materialization? 
This need to be answered after we annotate {\em Sp}, {\em Ind}, and {\em wt}. 
\end{itemize}

\item Requirement 4:  {\em Sp} with the same name should refer to the same species (e.g., a bird named {\em Aus} flies from sub-plot (1,A) to (1,B).)  
But the {\em Sp} with the same name at different sub-plot should refer to different observations of a specie.
%\begin{itemize}
%\item At the first glance, to represent this, {\em SpName} for {\em Sp} should be specified with ``key yes''. 
%It should have context {\em sub-plot} which is specified with ``identifying yes''. 
%\item {\bf The same problem as Case 3: The key measurements for the {\em Sp} observation is different from the key measurement  of the entity {\em Sp}}.
%the key measurements for the species observation are \{Plotlabel, SubPlotLabel, SpName\}. 
%However, the key measurements for the species entity is just \{SpName\}. 
%\end{itemize}

\item Requirement 5:  {\em Ind} with the same label and and the same species name should refer to the same species.
But the individual (with the same label and the same species name) at different sub-plot should refer to different species observations.
%\begin{itemize}
%\item {\bf The same problem as Case 4 and Case 5: The key measurements for the {\em Ind} observation is different from the key measurement  of the entity {\em Ind}.}
%the key measurements for the species observation are \{Plotlabel, SubPlotLabel, SpName, Ind\}. 
%However, the key measurement for the species entity is just \{SpName, Ind\}.  When two rows have the same value on these two columns, they represent the same entity instance. 
%In this case, the observation context of Ind is {\em Sp} and {\em Sub-plot}. But the entity context of Ind is just {\em Sp}. 
%\end{itemize} 
\end{itemize}

The annotation can be done as follows. 
\begin{itemize}
\item Observation type: Plot {\em distinct yes}
  \begin{itemize}
  \item Measurement type: PlotLabel {\em key yes}
  \end{itemize}
\item Observation type: SubPlot {\em distinct yes}
  \begin{itemize}
    \item Measurement type: SubPlotLabel {\em key yes}
      \item Context: Plot {\em identifying yes}. This means that the
        key measurement for SubPlot (SubPlotLabel) and the key
        measurement for Plot (PlotLabel) together form the key of
        SubPlot observation (and its related entity too.)
      \end{itemize}
\item Observation type: Treatment {\em distinct yes}
  \begin{itemize}
    \item EntityType: TreatmentProcedure 
    \item Measurement type: TmntType {\em key yes}
   \item Context: SubPlot {\em Identifying yes}
  \end{itemize}
\item Observation type: SpeciesIndividual
  \begin{itemize}
      \item Measurement type: Species, {\em key yes}
      \item Measurement type: IndLabel (represent {\em Individual}), {\em key yes} 
      \item Measurement type: Weight
      \item Context: SubPlot. Note: (1) if the users want to treat the
        individual (with the same lavel and the same species) at
        different sub-plots as different observation instances, here,
        {\em identifying yes} should be specified. Otherwise, (2) if the users want to treat the
        individual (with the same lavel and the same species) at
        different sub-plots as SAME observation instances, here, we
        don't need to specify {\em  identifying yes}.
   \end{itemize} 
\end{itemize}

%In summary, we can get a better idea about the problem described in
%the above use cases can when we answer the following two simple
%questions: \\
%Q1:  Will an entity type and an observation type (which is of the
%given entity type) always have the same key measurement type(s)?
%     The above use cases give situations that the answer is no.  \\
%Q2: is {\em identifying} itself enough to distinguish the key
%measurement for observation types and for entity types? 
%    My temporary answer to this question is no.

%A general thinking: 
%the counterpart in RDB (Relational DataBase) is a relational scheme with key attributes. 
%Here, we have two levels of objects: entity level and instance level.
%Then, for different levels of objects, we need to have different ways to specify their key measurements.

%\newpage
